---
title: "WGCNA - Busselton and Celtic Fire - Paper figures"
author: "Jonathan Ish-Horowicz"
date: "02/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preamble

Load packages

```{r}
# Import libraries
set.seed(12345)

library(WGCNA)
disableWGCNAThreads()
options(stringsAsFactors=FALSE)

library(phyloseq)
library(data.table)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(matrixStats)
library(psych)
library(stringr)
library(gplots)
library(foreach)
library(caret)
library(dendextend)
library(forcats)
library(compositions)

library(ggplot2)
theme_set(theme_bw(base_size=16))
library(ggpubr)
library(ggrepel)
library(ggforce)
library(drlib)
library(RColorBrewer)
```

Load data

```{r}
load("../data/Bus_CF_working_noduplicates.Rdata")
Bus_CF1
```

Extract OTU tables at each site

```{r}
X <- list()
X[["cf_lll"]] <- t(as.matrix(otu_table(subset_samples(Bus_CF1, SwabSite=="Brushing LLL 16S"))))
X[["cf_lul"]] <- t(as.matrix(otu_table(subset_samples(Bus_CF1, SwabSite=="Brushing LUL 16S"))))
X[["cf_ots"]] <- t(as.matrix(otu_table(subset_samples(Bus_CF1, SwabSite=="OTS"))))
X[["bus_ots"]] <- t(as.matrix(otu_table(subset_samples(Bus_CF1, SwabSite=="Throat"))))
lapply(X, dim)
```

Load the taxonomy

```{r}
taxonomy_df <- as.data.frame(phyloseq::tax_table(Bus_CF1)) %>%
  dplyr::rename(OTU=OTUID)
```

# WGCNA

## Preprocessing

### OTU filtering

Check which OTUs with zero-variance (they also have zero prevalence)

```{r}
zero_variance_otus <- lapply(X, function(x) colnames(x)[colVars(x)==0])
print(sapply(zero_variance_otus, length))
zero_variance_otus <- unique(do.call(c, zero_variance_otus))
cat(sprintf("%d OTUs have zero-variance in at least one subset\n", length(zero_variance_otus)))
# X <- lapply(X, function(x) x[,!colnames(x) %in% zero_variance_otus])
# lapply(X, dim)
```
Remove these OTUs from the OTU tables.

```{r}
PREV_THRESH <- 20.0
prevalence_by_site <- lapply(
  X,
  function(x) apply(x, 2, function(xx) 100.0*sum(xx!=0)/length(xx)) %>%
    as.data.frame() %>%
    rownames_to_column("OTU") %>%
    dplyr::rename(prevalence=".")) %>%
  rbindlist(idcol="site")

included_OTUs <- unique(prevalence_by_site$OTU[prevalence_by_site$prevalence>=20])
cat(sprintf("%d OTUs are above the prevalence threshold of %.1f%%\n", length(included_OTUs), PREV_THRESH))

# subset OTUs
X <- lapply(X, function(x) x[,colnames(x) %in% included_OTUs])
X <- lapply(X, function(x) x %>% as.data.frame() %>% as.matrix())
X <- lapply(X, function(x) x[,colVars(x)>0])
lapply(X, dim)
```

## Network Construction

### Adjacency network

Using Spearman correlation.

```{r}
powerVector <- 1:15
RsquaredCut <- 0.85
NETWORK_TYPE <- "unsigned"
corOptions <- list(method="spearman")

beta_choice_out <- lapply(
    X,
    function(x) pickSoftThreshold(x,
                                  RsquaredCut=RsquaredCut, 
                                  corFnc=cor,
                                  corOptions=corOptions,
                                  powerVector=powerVector,
                                  removeFirst=FALSE,
                                  blockSize=NULL,
                                  networkType=NETWORK_TYPE,
                                  moreNetworkConcepts=FALSE, 
                                  verbose=0, indent=0)
  )
beta <- sapply(beta_choice_out, function(x) x$powerEstimate)
beta
```

A nicer looking plot

```{r}
fitIndices <- rbindlist(lapply(beta_choice_out, function(x) x$fitIndices), idcol="site")

tmpdf <- data.frame(y=c(RsquaredCut, NA), name=c("Signed R^2", "Mean connectivity"))
tmpdf$name <- factor(tmpdf$name, levels=c("Signed R^2", "Mean connectivity"))
  
fitIndices %>%
  mutate(signed_Rsq=-sign(slope)*SFT.R.sq) %>%
  select(site, Power, signed_Rsq, mean.k.) %>%
  pivot_longer(c(signed_Rsq, mean.k.)) %>%
  mutate(name=recode(name, mean.k.="Mean connectivity", signed_Rsq="Signed R^2")) %>%
  filter(name=="Signed R^2") %>%
  mutate(name=factor(name, levels=c("Signed R^2", "Mean connectivity"))) %>%
  ggplot(aes(x=Power, y=value, colour=site)) +
  # facet_wrap(~site) +
  geom_line() + geom_point() +
  theme(text=element_text(size=20), legend.position="top", legend.title=element_blank()) +
  geom_hline(yintercept=RsquaredCut, colour="red") +
  ylab("Signed R^2") + xlab("Soft power")

fitIndices %>%
  mutate(signed_Rsq=-sign(slope)*SFT.R.sq) %>%
  select(site, Power, signed_Rsq, mean.k.) %>%
  pivot_longer(c(signed_Rsq, mean.k.)) %>%
  mutate(name=recode(name, mean.k.="Mean connectivity", signed_Rsq="Signed R^2")) %>%
  filter(name!="Signed R^2") %>%
  mutate(name=factor(name, levels=c("Signed R^2", "Mean connectivity"))) %>%
  ggplot(aes(x=Power, y=value, colour=site)) +
  geom_line() + geom_point() +
  theme(text=element_text(size=20), legend.position="top", legend.title=element_blank()) +
  ylab("Mean connectivity") + xlab("Soft power")
```

### Adjacency and TOM

Calculate adjancency and TOM

```{r}
X_logtrans <- lapply(X, log1p)
stopifnot(all(sapply(X_logtrans, function(x) sum(is.na(x))==0)))

# adjacency calculation
adj <- lapply(names(X),
              function(x) adjacency(X[[x]], type=NETWORK_TYPE, power=beta[[x]], corOptions=corOptions))
names(adj) <- names(X)

# TOM similarity
TOM <- lapply(adj, function(x) TOMsimilarity(x, TOMType=NETWORK_TYPE, verbose=0))
for(site in names(TOM)) {
  colnames(TOM[[site]]) <- colnames(X[[site]])
  rownames(TOM[[site]]) <- colnames(X[[site]]) 
}

# TOM dissimilarity
dTOM <- lapply(TOM, function(tom) 1.0 - tom)

TOM %>%
  saveRDS(file="../results/wgcna/tom_matrices.rds")
```

### Module detection

```{r}
minClusterSize  <- 10

otu_trees <- lapply(dTOM, function(dtom) hclust(as.dist(dtom), method="average"))

premerge_modules <- lapply(names(otu_trees),
                  function(site) cutreeDynamic(dendro=otu_trees[[site]],
                                               distM=dTOM[[site]],
                                               deepSplit=2, pamRespectsDendro=FALSE,
                                               minClusterSize=minClusterSize)
)
names(premerge_modules) <- names(otu_trees)
premerge_colours <- lapply(premerge_modules, labels2colors)
```

Merge similar modules

```{r}
MEDissThres <- 0.35

module_eigengene_out <- lapply(names(premerge_colours),
                               function(x) moduleEigengenes(X_logtrans[[x]],
                                                               colors=premerge_colours[[x]])
)
names(module_eigengene_out) <- names(premerge_colours)
module_eigengenes <- lapply(module_eigengene_out, function(x) x$eigengenes)

module_eigengne_trees <- lapply(module_eigengenes, function(me) hclust(as.dist(1-cor(me)), method="average"))

for(label in names(module_eigengne_trees)) {
  plot(module_eigengne_trees[[label]], main=label, xlab="", sub="")
    abline(h=MEDissThres, col="red")
}
```


Do the merging:

```{r}
merge_out <- lapply(names(premerge_colours),
                    function(x) mergeCloseModules(X_logtrans[[x]],
                                                  premerge_colours[[x]],
                                                  cutHeight=MEDissThres)
)
names(merge_out) <- names(premerge_colours)
mergedColors <- lapply(merge_out, function(x) x$colors)
mergedMEs <- lapply(merge_out, function(x) x$newMEs)
```

Plot the modules before and after the merge

```{r}
for(site in names(otu_trees)) {
  png(sprintf("../plots/wgcna/module_dendrograms_%s.png", site), height=1000, width=2000, res=300)
  plotDendroAndColors(otu_trees[[site]],
                        cbind(premerge_colours[[site]], mergedColors[[site]]),
                        c("Dynamic Tree Cut", "Merged dynamic"),
                        dendroLabels = FALSE, hang = 0.03,
                        addGuide = TRUE, guideHang = 0.05,
                        main=site,
                        )
  dev.off()
}
  
```

# Module membership

```{r}
calc.mod.mem <- function(dat, eig.genes, alpha=0.05, ...) {
  
  # computes module membership (correlation between log-transformed)
  # OTU counts and module eigengenes. Also returns corresponding p-value
  
  out <- corr.test(
    dat, eig.genes,
    alpha=alpha, method="pearson", adjust="none", minlength=100,
    ...
  )
  
  corr.vals <- as.data.frame(out$r)
  p.vals <- as.data.frame(out$p)
  ci <- as.data.frame(out$ci)
  
  corr.vals <- as.data.frame(out$r) %>%
    rownames_to_column("OTU") %>%
    pivot_longer(cols=-OTU, names_to="module", values_to="correlation")
  
  p.vals <- as.data.frame(out$p) %>%
    rownames_to_column("OTU") %>%
    pivot_longer(cols=-OTU, names_to="module", values_to="pvalue")
  
  df <- full_join(corr.vals, p.vals, by=c("OTU", "module"))
  return(df)
}

module_membership <- lapply(names(mergedMEs),
                            function(x) calc.mod.mem(X_logtrans[[x]], mergedMEs[[x]]))
names(module_membership) <- names(mergedMEs)

combined_mm <- rbindlist(module_membership, idcol="label") %>%
  mutate(qvalue=p.adjust(pvalue, method="fdr"))

module_membership %>%
  saveRDS(file="../results/wgcna/module_membership.rds")
```

<!-- Write to csv files in wide format -->

<!-- ```{r} -->
<!-- for(label_ in names(module_membership)) { -->

<!--   long_df <- combined_mm %>% -->
<!--     filter(label==label_) %>% -->
<!--     select(-label) -->
<!--   module_names <- unique(long_df$module) -->
<!--   column_order <- apply( -->
<!--     expand.grid(c("correlation_", "pvalue_", "qvalue_"), module_names), -->
<!--     1, -->
<!--     function(x) sprintf("%s%s", x[[1]], x[[2]])) -->
<!--   column_order <- c("OTU", column_order) -->

<!--   wide_df <- long_df %>% -->
<!--     pivot_wider(names_from=module, values_from=c(correlation, pvalue, qvalue)) %>% -->
<!--     select(all_of(column_order)) %>% -->
<!--     fwrite(file=sprintf("../csv/module_membership_%s.csv", label_), sep=",") -->
<!-- } -->
<!-- ``` -->

<!-- Save X as rds file -->

<!-- ```{r} -->
<!-- saveRDS(X, file="../saved_results/X_wgcna.rds") -->
<!-- ``` -->

# Module assignments

```{r}
module_assignments <- lapply(setNames(names(mergedColors), names(mergedColors)),
       function(x) data.frame(OTU=colnames(X[[x]]),
                                              module=mergedColors[[x]]))

module_assignments_long <-  module_assignments %>%
  rbindlist(idcol="label")
  
module_assignments_wide <- module_assignments_long %>%
  pivot_wider(names_from=label, values_from=module) 

module_assignments %>%
  saveRDS(file="../results/wgcna/module_assignments.rds")
```

# Hub OTUs

```{r}
# for merged modules
hub_otus <- lapply(
  names(mergedColors),
  function(x) chooseTopHubInEachModule(
    X_logtrans[[x]], mergedColors[[x]], power=beta[[x]], type=NETWORK_TYPE
  )
)
names(hub_otus) <- names(mergedColors)

hub_otu_labels <- lapply(hub_otus,
       function(x) setNames(c(paste0(names(x), " (", gsub('_[^_]*$', '', x), ")"), "grey"),
                            c(names(x), "grey"))) 
saveRDS(hub_otu_labels, file="../results/wgcna/hub_otu_labels.rds")

module_name_tbl <- lapply(hub_otu_labels,
                          function(x) x %>% as.data.frame %>% rownames_to_column("module")) %>%
  rbindlist(idcol="site") %>%
  rename(module_longname=".")

module_name_tbl %>%
  fwrite("../results/wgcna/wgcna_module_names.csv")
```

# Association with traits

Test for associations between module eigenvectors and traits.

```{r}
module_eigenvectors <- mergedMEs
module_eigenvectors %>%
  saveRDS("../results/wgcna/module_eigenvectors.rds")
```

Utility function definitions - calculate associations between module eigenvectors for a continuous (correlation) or discrete (Kruskal-Wallis) trait.

```{r}
library(effectsize)
library(broom)
library(corrr)

cor_student_test <- function(MEs, traits) {
  stopifnot(nrow(MEs)==nrow(traits))
  corrvals <- cor(MEs, traits, use="p")
  pvals <- corPvalueStudent(corrvals, nrow(MEs))
  
  out <- corrvals %>%
    as.data.frame() %>%
    rownames_to_column("module") %>%
    pivot_longer(-module, values_to="corr", names_to="trait") %>%
    inner_join(
      pvals %>%
        as.data.frame() %>%
        rownames_to_column("module") %>%
        pivot_longer(-module, values_to="p", names_to="trait"),
      by=c("module", "trait")
    )
  
  return(out)
}

rowwise_kw_test <- function(MEs, traits) {
  stopifnot(nrow(MEs)==nrow(traits))
  n_MEs <- ncol(MEs)
  n_traits <- ncol(traits)
  
  # p-values from KW test
  p_values <- matrix(NA, nrow=n_MEs, ncol=n_traits)
  for(i in 1:n_MEs) {
    for(j in 1:n_traits) {
      p_values[i,j] <- kruskal.test(MEs[,i], traits[,j])$p.value
    }
  }
  rownames(p_values) <- colnames(MEs)
  colnames(p_values) <- colnames(traits)
  
  # rank epsilon squared effect size
  eps_sq <- data.frame()
  for(i in 1:n_MEs) {
    for(j in 1:n_traits) {
      eps_sq <- rbind(
        eps_sq,
        data.frame(
          rank_epsilon_squared(MEs[,i], traits[,j]) %>% as.data.frame(),
          module=colnames(MEs)[[i]],
          trait=colnames(traits)[[j]]
        )
      )
    }
  }
  
  # combine into single dataframe
  out <- p_values %>%
    as.data.frame() %>%
    rownames_to_column("module") %>%
    pivot_longer(-module, values_to="p", names_to="trait") %>%
    inner_join(
      eps_sq,
      by=c("module", "trait")
    )
  return(out)
}
```

Read formatted metadata - produced using `markdown/metadata_formatting.Rmd`

```{r}
metadata <- fread("../data/metadata/formatted_metadata.csv") %>%
  as.data.frame()
```

Check which samples we have

```{r}
for(site in names(X)) {
  cat(site, ":\n")
  samples_with_metadata <- rownames(X[[site]])[rownames(X[[site]]) %in% metadata$SampleID]
  cat(length(samples_with_metadata), "of", nrow(X[[site]]), "samples in metadata\n")
}
```

Count the number of non-NA samples for each trait at each site.

TODO: check how GERD and diabetes patients are handled

```{r}
discrete_phenotype_names <- c("Asthma", "Smoking", "Sex", "Severity")
continuous_phenotype_names <- c("BMI", "SmokingQuit", "Age", "WCC", "pc_Neutrophils", "pc_Eosinophils")

y_discrete <- metadata[,discrete_phenotype_names] %>% as.matrix()
y_continuous <- metadata[,continuous_phenotype_names] %>% as.matrix()

rownames(y_continuous) <- metadata$SampleID
rownames(y_discrete) <- metadata$SampleID

frac_valid_samples <- list()

for(label in names(module_eigenvectors)) {
  cat(label, "\n")

  # continous trait
  id <- paste0("continuous.", label)
  included_ids <- intersect(rownames(module_eigenvectors[[label]]),
                            rownames(y_continuous))

  frac_valid_samples[[id]] <- apply(y_continuous[included_ids,],
                                 2,
                                 function(x) sum(!is.na(x))/length(x))
  
  # discrete traits
  id <- paste0("discrete.", label)
  included_ids <- intersect(rownames(module_eigenvectors[[label]]),
                            rownames(y_discrete))
  frac_valid_samples[[id]] <- apply(y_discrete[included_ids,],
                                 2,
                                 function(x) sum(!is.na(x))/length(x))
}

frac_valid_samples
```

Don't associate SmokingQuit with cf.lll samples - only 10/111 samples are not NA.

Calculate the associations.

```{r}
module_trait_assocs <- list()

for(label in names(module_eigenvectors)) {
  cat(label, "\n")
  site <- label

  #
  # calculate associations
  
  # continuous traits
  id <- paste0("continuous.", label)
  included_ids <- intersect(rownames(module_eigenvectors[[label]]),
                            rownames(y_continuous))
  
  if(length(included_ids)!=nrow(module_eigenvectors[[label]]))
    warning(sprintf("Missing &d continuous samples for &s\n", nrow(module_eigenvectors)-length(included_ids), label))
  
  included_traits <- n_valid_samples[[paste0("continuous.", label)]]
  included_traits <- names(included_traits)[included_traits>0.5]

  module_trait_assocs[[id]] <- cor_student_test(module_eigenvectors[[label]][included_ids,],
                                                y_continuous[included_ids,included_traits])
  
  # discrete traits
  id <- paste0("discrete.", label)
  included_ids <- intersect(rownames(module_eigenvectors[[label]]),
                            rownames(y_discrete))
  if(length(included_ids)!=nrow(module_eigenvectors[[label]]))
    warning(sprintf("Missing &d discrete samples for &s\n", nrow(module_eigenvectors)-length(included_ids), label))
  
  included_traits <- n_valid_samples[[paste0("discrete.", label)]]
  included_traits <- names(included_traits)[included_traits>0.5]

  module_trait_assocs[[id]] <- rowwise_kw_test(module_eigenvectors[[label]][included_ids,],
                                               y_discrete[included_ids,included_traits])
}
```
Do the multiple testing correction (FDR)

```{r}
q_values <- lapply(module_trait_assocs, function(x) x %>% select(module, trait, p)) %>%
  rbindlist(idcol="label") %>%
  mutate(q=p.adjust(p, method="fdr")) %>%
  arrange(q)

module_trait_assocs_corrected <- list()
for(label_ in names(module_trait_assocs)) {
  module_trait_assocs_corrected[[ label_ ]] <- module_trait_assocs[[ label_ ]] %>%
    inner_join(
      q_values %>%
        filter(label==label_) %>%
        select(-label),
      by=c("module", "trait", "p")
    )
  
}
```

Save the results

```{r}
saveRDS(module_trait_assocs_corrected, "../results/wgcna/module_trait_assocs.rds")
```
